---
title: "Supplementary Material: Assessing the validity of a calcifying oral biofilm model as a suitable proxy for dental calculus"
author: "B.P.Bartholdy, I.M.Velsko, S.Gur-Arieh, Z.Fagernäs, A.G.Henry"
crossref: 
  #fig-prefix: S
  fig-title: Figure S
  #fig-labels: roman
  #labels: arabic
format:
  html: 
    toc: true
    code-fold: true
    output-file: "supplementary-material.html"
  pdf: 
    toc: true
    output-file: "supplementary-material.pdf"
    fig-pos: "H"
    knitr: 
      opts_chunk: 
        echo: false
execute: 
  warning: false
knitr: 
  opts_chunk: 
    collapse: true
    warning: false
    fig-path: "figures/supp-"
prefer-html: true
editor_options: 
  chunk_output_type: console
---

```{r}
#| label: setup
#| include: false
library(here)
library(dplyr)
library(tidyr)
library(forcats)
library(readr)
library(stringr)
library(ggplot2)
library(plotly)
library(cuperdec)
knitr::opts_knit$set(root.dir = here())
#knitr::opts_chunk$set(fig.path = "figures/supp-")

# upload ftir data

ftir_metadata <- read_tsv(here("01-documentation/ftir-metadata.tsv"))
ftir_data_raw <- read_csv(here("05-results/ftir-data.csv"))
ftir_grind_data <- read_csv(here("05-results/grind-data_cleaned.csv"))

# upload dna data

dna_metadata <- read_tsv(here("01-documentation/metadata.tsv"))
dna_analysis_metadata <- read_tsv(here("01-documentation/analysis-metadata.tsv"))
dna_experiment_metadata <- read_tsv(here("01-documentation/experiment-metadata.tsv"))
otu_table <- read_tsv(here("05-results/post-decontam_taxatable.tsv"))
bac_properties <- read_tsv(here("01-documentation/species-properties.tsv"))
genus_oxytol <- read_tsv(here("01-documentation/genus-O2tolerance.tsv"))
otu_decontam <- read_tsv(here("05-results/post-decontam_taxatable.tsv"))
load(here("05-results/spca_byoc.rda"))
load(here("05-results/spca_species.rda"))
alpha_div <- read_tsv(here("05-results/alpha-diversity.tsv"))
byoc_logf_full <- read_tsv(here("05-results/byoc_logf-full.tsv"))
plaque_logf_full <- read_tsv(here("05-results/plaque_logf-full.tsv"))
sourcetracker2 <- read_tsv(here("04-analysis/sourcetracker/sourcetracker2_output/mixing_proportions.txt"))
all_data_long <- read_tsv(here("04-analysis/sourcetracker/source-comb_long.tsv"))
clr_compar_long <- readr::read_tsv(here("05-results/clr-compar.tsv")) %>%
  pivot_longer(-sample, values_to = "clr_count", names_to = "species")
clr_byoc_long <- readr::read_tsv(here("05-results/clr-byoc.tsv")) %>%
  pivot_longer(-sample, values_to = "clr_count", names_to = "species")
pca_loadings <- readr::read_tsv(here("05-results/all-pca-loadings.tsv"))
#load(here("05-results/spca_byoc.rda"))
#load(here("05-results/spca_species.rda"))

source(here("02-scripts/functions.R"))
source(here("02-scripts/99_setup-qmd.R"))
source(here("02-scripts/99_ftir-figures.R"))
source(here("02-scripts/99_metagen-figures.R"))

```

## Samples

Samples taken for DNA sequencing and FTIR analysis. Samples for DNA were collected
from a seperate experimental run than samples for DNA, but experimental conditions
were the same in each. Samples for DNA were medium collected from the bottom of
each well after three days of growth (before medium was refreshed). Samples for
FTIR were taken directly from the biofilm and dried prior to analysis.

### DNA

```{r}
#| label: tbl-byoc-samples
#| tbl-cap: "Table of biofilm samples from this study. Includes which day in the experiment the sample was taken, and sample type (Env)."
dna_experiment_metadata %>% 
  #filter(str_detect(`#SampleID`, "LIB|SYN", negate = F)) %>% 
  select(c(`#SampleID`, day, Env)) %>% 
  arrange(day, Env) %>% 
  knitr::kable()
```

### FTIR

```{r}
#| label: tbl-ftir-samples
#| tbl-cap: "Table of oral reference samples. Includes which day in the experiment the sample was taken, and sample type."
ftir_metadata %>% 
  #filter(str_detect(sample_id, "", negate = T)) %>% 
  select(c(sample_id, day, sample_type)) %>% 
  arrange(day) %>% 
  knitr::kable()
```


## Reference database and sequences

The reference database used in the EAGER pipeline was the Standard Kraken 2 database,
downloaded from <https://genome-idx.s3.amazonaws.com/kraken/k2_standard_20210517.tar.gz>

```sh
wget https://genome-idx.s3.amazonaws.com/kraken/k2_standard_20210517.tar.gz
```

The human reference genome GRCh38 was downloaded on 2022-04-22.

```sh
wget ftp://ftp.ncbi.nlm.nih.gov/refseq/H_sapiens/annotation/GRCh38_latest/refseq_identifiers/GRCh38_latest_genomic.fna.gz
```

```{r}
#| label: tbl-oral-samples
#| tbl-cap: "Table of oral reference samples. Includes sample type (Env), and associated project and study with DOI."
dna_metadata %>% 
  filter(
   str_detect(`#SampleID`, "LIB|SYN", negate = T),
    !Env %in% env_controls
  ) %>% 
  select(!c(SourceSink, download_link, Study)) %>% 
  arrange(Env) %>% 
  knitr::kable()
```

```{r}
#| label: tbl-env-samples
#| tbl-cap: "Table of environmental reference samples (contamination testing). Includes sample type (Env), and associated project and study with DOI."
dna_metadata %>% 
  filter(
    str_detect(`#SampleID`, "LIB|SYN", negate = T),
    Env %in% env_controls
  ) %>% 
  select(!c(SourceSink, download_link, Study)) %>% 
  arrange(Env) %>% 
  knitr::kable()
```

## Pre-processing

### SourceTracker2

Steps taken for SourceTracker analysis.

OTU table was filtered for relative abundance. Percent abundance of each taxon
across all samples was calculated and then taxa with lower than 0.001% abundance
were filtered out.

Installation of Qiime2 and dev version of SourceTracker via conda:

```sh
wget https://data.qiime2.org/distro/core/qiime2-2022.2-py38-linux-conda.yml
conda env create -n qiime2-2022.2 --file qiime2-2022.2-py38-linux-conda.yml
rm qiime2-2022.2-py38-linux-conda.yml # cleanup
conda activate qiime2-2022.2
pip install https://github.com/biota/sourcetracker2/archive/master.zip
```

Installation of Qiime1 to access filter_samples_from_otu_table.py

```sh
conda create -n qiime1 python=2.7 qiime -c bioconda
```

Convert OTU table from *.tsv* to *.biom*.

```sh
biom convert -i 04-analysis/OTUfilter_table.tsv -o 04-analysis/sourcetracker/OTUfilter-table-from-tsv_json.biom --table-type="OTU table" --to-json
```

Filter OTU table

```sh
filter_samples_from_otu_table.py \
-i 04-analysis/sourcetracker/OTUfilter-table-from-tsv_json.biom \
-o 04-analysis/sourcetracker/OTUs1000filter_table.biom \
-n 1000
```

Table summary

```sh
biom summarize-table -i 04-analysis/sourcetracker/OTUs1000filter_table.biom > 04-analysis/sourcetracker/summary_OTUs1000filter-table.txt
```

Convert to TSV for use with the **decontam** package.

```sh
biom convert -i 04-analysis/sourcetracker/OTUs1000filter_table.biom -o 04-analysis/decontam/pre-decontam_OTUfiltered-table_from-biom.tsv --to-tsv
```

Run SourceTracker2 on the filtered OTU table with rarefaction depth of 1000
for both source and samples. Samples and sources were mapped in the
[ST_comb-plaque-map.txt](../04-analysis/sourcetracker/ST_comb-plaque-map.txt).
The plaque source is a combination of supragingival and subgingival plaque.

```sh
conda activate qiime2-2022.2
```

```sh
sourcetracker2 \
    -i 04-analysis/sourcetracker/OTUs1000filter_table.biom \
    -m 04-analysis/sourcetracker/ST_comb-plaque-map.txt \
    --source_sink_column SourceSink  \
    --source_column_value source \
    --source_rarefaction_depth 1000 \
    --sink_rarefaction_depth 1000 \
    --sink_column_value sink \
    --source_category_column Env \
    -o 04-analysis/sourcetracker/sourcetracker2_output \
    --jobs 2 \
    --per_sink_feature_assignments
```

Plots were created of estimated contributions of various sources to the saliva,
model calculus and medium samples. Samples are arranged from left to right by how
late in the experiment they were sampled, with left being the earliest samples
(@fig-sourcetracker2). The output from SourceTracker2 was compared to a dataset
of known oral species (@fig-oral-sources).

```{r}
#| label: st2-plot-setup
#| echo: false
sourcetracker2_long <- sourcetracker2_longer()
day_order <- dna_experiment_metadata %>%
  mutate(Env = factor(
    Env, levels = c("saliva", "medium", "byoc_calculus") # force level order so it doesn't order alphabetically
    ) 
  ) %>% 
  group_by(Env) %>%
  arrange(day, .by_group = T) %>%
  filter(`#SampleID` %in% colnames(sourcetracker2)) %>%
  mutate(rm = if_else(`#SampleID` %in% dna_analysis_metadata$`#SampleID`, F, T),
         col = case_when(rm == T ~ "red", # excluded samples coloured red
                         rm == F ~ "black"))
```


```{r}
#| label: fig-sourcetracker2
#| fig-cap: "Estimated proportion of source composition of the abundance-filtered oral biofilm model samples using SourceTracker2. Names of removed samples in red text."
sourcetracker2_long %>% 
  ggplot(aes(y = SampleID, x= proportion, fill = source)) +
    geom_col() +
    theme_minimal() +
    scale_y_discrete(limits = day_order$`#SampleID`) +
    scale_fill_viridis_d(option = "C") +
    theme(axis.title = element_blank(),
          axis.text.y = element_text(vjust = 0.5, hjust = 0.5, colour = day_order$col))
```

```{r}
#| label: fig-oral-sources
#| fig-cap: "Estimated proportion of oral or other composition of the abundance-filtered oral biofilm model samples using the oral database from the cuperdec R package. Names of removed samples in red text."
oral_taxa <- cuperdec_database_ex %>%
  filter(isolation_source == "oral")
all_data_long %>% 
  filter(count > 0) %>% 
  mutate(oral_source = if_else(taxon %in% oral_taxa$species, "oral", "other")) %>%
  ggplot(aes(y = SampleID, x = count, fill = oral_source)) +
    geom_col(position = "fill") +
    scale_fill_viridis_d(option = "C") +
    theme_minimal() +
    theme(axis.text.y = element_text(colour = day_order$col),
          axis.title.y = element_blank()) +
    scale_y_discrete(limits = day_order$`#SampleID`)
```

### decontam

```{r}
species_table_long <- otu_decontam %>%
  pivot_longer(cols = where(is.numeric), names_to = "sample", values_to = "count")

byoc_table_long <- species_table_long %>% 
  filter(str_detect(sample, "SYN"),
         sample %in% dna_analysis_metadata$`#SampleID`,
         count > 0)

species_summ <- byoc_table_long %>%
  group_by(sample) %>%
  count(`#OTU ID`) %>%
  summarise(count = sum(n))
```

After running `isContaminant()` and filtering out (non-oral) potential contaminants,
the samples contained between
`r paste(range(species_summ$count), collapse = " and ")`
species with a mean of
`r paste(round(mean(species_summ$count), 2))` (@fig-species-counts).

```{r}
#| label: fig-species-counts
#| fig-cap: "Species counts after removal of contaminants for biofilm model samples."

byoc_table_long %>% 
  group_by(sample) %>%
  count(`#OTU ID`) %>% 
  ggplot(aes(y = sample, x = n), fill = viridisLite::magma(1, begin = 0.5)) +
    geom_col() +
  theme_minimal() +
    theme(
      axis.line.x = element_line(),
      axis.title.x = element_blank(),
      panel.grid.major.y = element_blank(),
      axis.title.y = element_blank())
```


## Species composition

Counts from model biofilm and oral reference samples were transformed with a
centered ratio log-transform and ordered by PC1 loading
(@fig-species-pos1 & @fig-species-neg1).

```{r}
#| label: fig-species-pos1
#| fig-cap: "Centered ratio log-transform abundance of the top 200 species with the highest positive loading on PC1."
#| fig-height: 12
#| fig-width: 9
# top positive loadings on PC1
species_pos_pc1 +
    scale_fill_viridis_c() +
    theme(
      axis.title.y = element_blank(),
      axis.title.x = element_blank(),
      axis.text.x = element_text(
        angle = 90, size = 8, vjust = 0.5, hjust = 1
        ),
      axis.text.y = element_text(face = "italic")
      )
```

```{r}
#| label: fig-species-neg1
#| fig-cap: "Centered ratio log-transform abundance of the top 200 species with the highest negative loading on PC1."
#| fig-height: 12
#| fig-width: 9
# Top negative loadings on PC1
species_neg_pc1 +
  scale_fill_viridis_c() +
  theme(
    axis.title.y = element_blank(),
    axis.title.x = element_blank(),
    axis.text.x = element_text(
      angle = 90, size = 8, vjust = 0.5, hjust = 1
    ),
    axis.text.y = element_text(face = "italic")
  )
```


## FTIR spectra

Select FTIR spectra not shown in the main manuscript (@fig-ftir-spectra).

```{r}
#| label: fig-ftir-spectra
#| layout-ncol: 2
#| fig-height: 2.3
#| fig-cap: "FTIR spectra"
#| fig-subcap: 
#|   - "Day 7"
#|   - "Day 7"
#|   - "Day 12"
#|   - "Day 12"
ftir_spect_plot(ftir_data, "F7.2D1")
ftir_spect_plot(ftir_data, "F7.1A6")
ftir_spect_plot(ftir_data, "F12.1A5+F12.B1")
ftir_spect_plot(ftir_data, "F12.1D1+F12.1D2")


ftir_spect_plot(ftir_data, "F16.1B2")
ftir_spect_plot(ftir_data, "F16.1D6")
ftir_spect_plot(ftir_data, "F20.1A1")
ftir_spect_plot(ftir_data, "F20.1B5")
ftir_spect_plot(ftir_data, "F24.1B3")
ftir_spect_plot(ftir_data, "F24.2A4")
```

::: {.content-visible when-format="pdf"}
Interactive plots are available in the HTML output file.
:::

::: {.content-visible when-format="html:js"}

Interactive plots to explore FTIR spectra.

```{r eval=!knitr::is_latex_output()}
#| label: fig-ftir-int7
ftir_spect_plot(ftir_data, "F7.1A6_b") %>% 
  ggplotly()
```

```{r eval=!knitr::is_latex_output()}
#| label: fig-ftir-int12
ftir_spect_plot(ftir_data, "F12.1A5+F12.B1_B") %>% 
  ggplotly()
```

```{r eval=!knitr::is_latex_output()}
#| label: fig-ftir-int16
 ftir_spect_plot(ftir_data, "F16.1C6") %>% 
  ggplotly()
```

```{r eval=!knitr::is_latex_output()}
#| label: fig-ftir-int24
ftir_spect_plot(ftir_data, "F24.1A3") %>% 
  ggplotly()
```

:::

<!-- host interactive plot somewhere online? -->

## Software versions

### EAGER

Software versions:

```{r}
readr::read_tsv(here("01-documentation/software_versions.csv"), col_names = c("Software", "Version")) %>%
  knitr::kable()
```

### R

#### Session

```{r}
print(sessionInfo(), locale = F)
```

#### Packages

```{r}
#| label: tbl-list-packages
#| tbl-cap: "List of R packages and dependencies."
renv_packages_list <- jsonlite::read_json(here("renv.lock"))
do.call(rbind, renv_packages_list$Packages) %>% 
  as.data.frame() %>%
  select(Package, Version, Source) %>%
  knitr::kable(row.names = F)
```

