## Metagenomic classification

<!-- metagenomic analysis -->

<!-- SourceTracker -->

<!-- species diversity -->

## FTIR

```{r}
#| label: results-setup
#| include: false
grind_sample_order <- c( # make sure the grind samples are ordered correctly in plots
  "Archaeological calculus",
  "Artificial calculus day 16",
  "Artificial calculus day 20",
  "Artificial calculus day 24",
  "Archaeological bone",
  "Bone-Dentine",
  "Bone-Dentine_2",
  "Enamel",
  "Enamel_2",
  "Enamel_3")

# Preparation of grind curve plots
grind_all_plot <- ftir_grind_data %>%
  mutate(
    Sample_day = factor(Sample_day, levels = grind_sample_order)) %>% # reorder sampling day levels
  group_by(day, Sample) %>%
    ggplot(aes(x = FWHM, y = IRSF, col = Sample_day, shape = Sample_day)) +
    geom_point(size = 2, alpha = 0.6) +
    geom_smooth(method = "lm", se = F) +
    theme_minimal() +
    labs(col = "Sample", shape = "Sample",
         x = "FWHM of the 1035 peak",
         y = "Splitting factor")
```

```{r}
#| label: fig-experiment-spectra
#| fig-cap: "Select spectra from all experiment sampling days."
#| layout-ncol: 2
#| fig-subcap:
#|  - "A"
#|  - "B"
#|  - "C"
#|  - "D"
# day 7
ftir_data %>%
  filter(
    day == 7,
    sample != "F7",
    sample != "F7.1A6_b"
    ) %>%
  ggplot(aes(x = wavenumber, y = abs, col = sample)) +
    geom_line() +
    theme_classic() +
    scale_x_reverse()

# day 12
ftir_data %>%
  filter(
    day == 12
    #sample != "F7",
    #sample != "F7.1A6_b"
    ) %>%
  ggplot(aes(x = wavenumber, y = abs, col = sample)) +
    geom_line() +
    theme_classic() +
    scale_x_reverse()

# day 16
ftir_data %>%
  filter(
    day == 16,
    is.na(grind),
    sample != "F16.2D2",
    sample != "F16.1D6",
    sample != "F16.1B2"
    ) %>%
  ggplot(aes(x = wavenumber, y = abs, col = sample)) +
    geom_line() +
    theme_classic() +
    scale_x_reverse()

# day 24 (final product)
ftir_data %>%
  filter(
    day == 24,
    is.na(grind),
    str_detect(sample, "[_]", negate = T),
    sample != "F24.1B3",
    sample != "F24.2C3",
    sample != "F24.2D3",
    sample != "F24.2A4",
    ) %>%
  ggplot(aes(x = wavenumber, y = abs, col = sample)) +
    geom_line() +
    theme_classic() +
    scale_x_reverse()
```

Day 7 spectra (#fig-experiment-spectra) have large O--H and amide A absorbance
bands in stretching mode around 3400 cm$^{-1}$,
as well as three marked CH~3~ and CH$_2$ stretching vibrations at 2960,
2920, and 2850 cm$^{-1}$. There is a clear amide I peak at 1650 and a less
pronounced amide II peak at 1545 cm$^{-1}$. 
In the 'fingerprint' region, C--O$_3^{2-}$ at 1400 absorbance band corresponding
to the v~3~ stretching vibrations, P--O~4~ absorbance band
corresponding to the v~3~ antisymmetric stretching vibrations
at 1080 cm$^{-1}$, and minor phosphate absorption bands around 500 cm$^{-1}$ in
sample F7.1A6, but absent in sample F7.2D1. The absorption bands at 1080
cm$^{-1}$ could also be a C--N stretching mode from aliphatic amines.
The relative absorbance of O--H and Amide I and II bands are higher than the
phosphate bands, representing a relatively higher content of lipids and proteins
than inorganic content. Large variation between spectra.

Day 12, amide I and II continue to be the dominant peaks, and a higher ratio
of both amide and O--H to PO~4~ v~3~ absorbance bands.
Three marked CH~3~ and CH$_2$ stretching vibrations at 2960,
2920, and 2850 cm$^{-1}$. Reduced variation between two of the three spectra.

Day 16, the ratio of O--H and amides to PO~4~ has shifted, with
the main peak shifting to the PO~4~ v~3~ absorbance band at 1040
(except in sample F16.1A2). A well-defined PO~4~ doublet at 600 and 560 is
present. Small CO$_3^{2-}$ stretching vibrations at 870 cm$^{-1}.
Decreased variability between the spectra, with most spectra exhibiting a
higher phosphate-to-protein/lipid ratio.

Day 24, large O--H and amide A absorbance bands in stretching mode around 3400 cm$^{-1}$,
as well as three minor CH~3~ and CH$_2$ stretching vibrations at 2960,
2920, and 2850 cm$^{-1}$.
Main peak of spectra is PO~4~ v~3~ at 1040
cm$^{-1}$, well-defined PO~4~ doublet at 600-550 cm$^{-1}$. Amide I band,
with small amide II and III bands. Carbonate peaks also present.
Very little variation between all of the spectra.

The main difference in the samples across the experiment, with
increasing age, is an
increase in the main peak at around 1040 cm-1, appearance of
the phosphate doublet around 600 and 560 cm-1 between day 12
and 16, an increase
in the ratio of the Amide I/II to peaks at around 1650 and 1540 cm-1,
and a reduction of the C--H2 and C--H3 stretching vibrations.

<!--move to discussion? -->
High lipid and protein content consistent with the presence of extracellular
polysaccharides and bacteria within a matrix. Microbial DNA and RNA may be
visible from peaks around 1200--800 cm$^{-1}$ on days 7 and 12, which are
later obscured by the increasing phosphate peak at 1040 cm$^{-1}$
[@jainIsolationCharacterization2013]. The presence of water indicated by
the O--H stretch is also consistent with a biofilm, which is around 90%
water [@bergerOralBiofilms2018]. As the samples mature, the ratio of proteins
and lipids to phosphates shift from predominance of organic content to
inorganic content.

Early spectra, days 7 and 12, are also similar to collagen spectra, with the
OH absorbance band at around 3400 cm$^{-1}$, amide I, II, and III peaks at
1659, 1552, and 1240 cm$^{-1}$, respectively 
[@rohanizadehUltrastructuralStudy2005; @martinezcortizasLinkingStructural2020].

The biofilm samples from the end of the experiment are similar to both reference
samples. The main
difference is a lower organic component in reference samples seen as a reduced
amide I peak at around 1637 compared to the carbonate peak at around 1420, and
an absence of amide II and III. Also a reduction in CH~3~ bands at 3000-2900 cm$^{-1}$.
<!-- FTIR spectra over course of experiment -->

<!-- grind curves -->

### Grind curves

```{r message=FALSE, warning=FALSE}
#| label: fig-grind-curve-inset
#| fig-cap: "(A) Grinding curves of multiple materials; and (B) calculus-only materials, including biofilm samples from three days, and an archaeological calclulus sample."
# produce plot with grind curve for all samples
grind_all_plot <- ftir_grind_data %>%
  mutate(
    Sample_day = factor(Sample_day, levels = grind_sample_order)) %>%
  group_by(day, Sample) %>%
    ggplot(aes(x = FWHM, y = IRSF, col = Sample_day, shape = Sample_day)) +
    geom_point(size = 2, alpha = 0.6) +
    geom_smooth(method = "lm", se = F) +
    theme_minimal() +
    labs(col = "Sample", shape = "Sample",
         x = "FWHM of the 1035 peak",
         y = "Splitting factor")

# isolate calculus samples to see diffs between days and the 'real deal'
grind_calc_plot <- ftir_grind_data %>%
  filter(Sample == "Artificial calculus" | Sample == "Archaeological calculus") %>%
  ggplot(aes(x = FWHM, y = IRSF, col = Sample_day, shape = Sample_day)) +
    geom_point(size = 2, alpha = 0.6) +
    geom_smooth(method = "lm", se = F) +
    labs(col = "Sample", shape = "Sample",
         x = "FWHM of the 1035 peak",
         y = "Splitting factor") +
    theme_minimal() +
    theme(
      axis.text = element_blank(),
      axis.title = element_blank(),
      axis.ticks = element_blank(),
      legend.position = "none",
      panel.border = element_rect(
        colour = "grey", 
        fill = "transparent", size = 1),
      panel.background = element_rect(fill = "white")
      ) +
    scale_colour_viridis_d(end = 0.4)

grind_all_plot + inset_element(grind_calc_plot, left = 0.5, bottom = 0.5, right = 1, top = 1) + 
  plot_layout(guides = "collect") + plot_annotation(tag_levels = "A")
```

Splitting factors were calculated from the grinding measurements, and a
grinding curve was created to explore crystallinity (crystal size) order and
disorder. Disorder is a steep slope and large FWHM.
Samples are similar to fresh bone and dentin. No appreciable differences between
days 16, 20, and 24.
<!-- FTIR spectra comparison of artificial and reference dental calculus -->