---
title: "Community composition"
output: html_notebook
---

```{r setup, include=FALSE}
library(tidyverse)
library(here)
species_abundance_long <- readr::read_tsv(here("05-results/species-abundance_long.tsv"))
genus_abundance_long <- readr::read_tsv(here("05-results/genus-abundance_long.tsv"))
metadata <- readr::read_tsv(here("01-documentation/metadata.tsv"))
bac_properties <- readr::read_tsv(here("01-documentation/species-properties.tsv"))

analysis_metadata <- metadata %>%
  filter(`#SampleID` %in% unique(species_abundance_long$sample))

species_properties <- bac_properties %>%
  right_join(species_abundance_long, by = "species") %>%
  mutate(abs = if_else(is.na(abs), FALSE, abs))

# top_species_names <- species_abundance_long %>% 
#   group_by(sample) %>% 
#   arrange(desc(rel_abund)) %>% 
#   slice_head(n = 4) %>% 
#   .$species %>% 
#   unique()

top_species_names <- species_abundance_long %>% 
  filter(rel_abund > 0.1) %>% 
  .$species %>% 
  unique()

top_abundance_species <- species_abundance_long %>% 
  mutate(
    top_species = case_when(
      rel_abund > 0.1 ~ species,
      TRUE ~ "other"
    ),
    # top_species = case_when(
    #  species %in% top_species_names ~ species,
    #   TRUE ~ "other"), # label all species below top 10 as 'other' so the rel_abund still sums to 1
    top_species = as_factor(top_species)) %>% 
  mutate(
    top_species = factor(
      top_species, 
      levels = c(levels(top_species)[-2], "other")
      ) # rearrange level 'other' to back
    ) 

top_genus_names <- genus_abundance_long %>% 
  group_by(sample) %>% 
  arrange(desc(rel_abund)) %>% 
  slice_head(n = 4) %>% 
  .$genus %>% 
  unique()

top_abundance_genus <- genus_abundance_long %>% 
  mutate(
    top_genus = case_when(
      genus %in% top_genus_names ~ genus,
      TRUE ~ "other"), # label all species below top 10 as 'other' so the rel_abund still sums to 1
    top_genus = as_factor(top_genus),
    top_genus = factor(top_genus, levels = c(levels(top_genus)[-2], "other")) # rearrange level 'other' to back
    ) 

# object to order plots by sampling day
day_order <- analysis_metadata %>% 
  filter(str_detect(`#SampleID`, "SYN")) %>% # only include samples from this study
  arrange(desc(day)) %>% # order by sampling day (ascending)
  .$`#SampleID` # isolate sample names


my_pallette2 <- c(viridisLite::magma(n = length(top_genus_names)), "grey") 
names(my_pallette2) <- levels(top_abundance_genus$top_genus)
my_scale2 <- scale_fill_manual(name = "top_genus", values = my_pallette2)

treatment_colours <- viridisLite::inferno(n = length(unique(analysis_metadata$treatment))) # define pallette and number of colours
env_colours <- viridisLite::magma(n = length(unique(analysis_metadata$Env))) # define pallette and number of colours

analysis_colours <- analysis_metadata %>%
  mutate(Env = as_factor(Env),
         treat_col = case_when(
           treatment == "potato" ~ treatment_colours[1],
           treatment == "wheat" ~ treatment_colours[2],
           treatment == "mix" ~ treatment_colours[3],
           treatment == "control" ~ treatment_colours[4],
           is.na(treatment) ~ "grey"),
         env_col = case_when(
           Env == "calculus" ~ env_colours[1],
           Env == "subgingival_plaque" ~ env_colours[2],
           Env == "supragingival_plaque" ~ env_colours[3],
           Env == "modern_calculus" ~ env_colours[4]
         ))
```

# Within the experiment

Plotting species abundances across the experiment.

## Genus composition

```{r}
top_abundance_genus %>%
  filter(str_detect(sample, "SYN")) %>% 
  ggplot(aes(x = rel_abund, y = sample, fill = top_genus)) +
    geom_col() +
    theme_void() +
    scale_y_discrete(limits = day_order) +
    my_scale2 +
    theme(axis.text = element_text())

genus_abundance_long %>%
  filter(str_detect(sample, "SYN")) %>% 
  mutate(top_genus = case_when(rel_abund > 0.05 ~ genus,
                               TRUE ~ "other")) %>% 
  ggplot(aes(x = rel_abund, y = sample, fill = top_genus)) +
    geom_col() +
    theme_void() +
    scale_y_discrete(limits = day_order) +
    #my_scale2 +
    theme(axis.text = element_text())
```



## Species composition

Only including the name of the top 10 species (relative abundance) within each
sample. Remaining species are designated as 'other'.


```{r}
 # label all species below top 10 as 'other' so the rel_abund still sums to 1 for plotting


# Plot colours
my_pallette <- c(viridisLite::plasma(n = length(top_species_names)), "grey") 
names(my_pallette) <- levels(top_abundance_species$top_species)
my_scale <- scale_fill_manual(name = "top_species", values = my_pallette)

# Plot colours
species_colours_sample <- top_abundance_species %>% 
  filter(str_detect(sample, "SYN")) %>%
  mutate(top_species = as.character(top_species)) %>% 
  .$top_species %>%
  unique()
my_pallette <- c(viridisLite::plasma(n = length(species_colours_sample) - 1), "grey") 
names(my_pallette) <- c(species_colours_sample[which(species_colours_sample != "other")], "other")
my_scale <- scale_fill_manual(name = "top_species", values = my_pallette)

top_abundance_species %>%
  filter(str_detect(sample, "SYN")) %>%
  #arrange(day) %>%
  ggplot(aes(x = rel_abund, y = sample, fill = top_species)) +
    geom_col() +
    theme_void() +
    scale_y_discrete(limits = day_order) +
    my_scale +
    theme(axis.text = element_text())

species_abundance_long %>%
  filter(str_detect(sample, "SYN")) %>% 
  mutate(top_species = case_when(rel_abund > 0.05 ~ species,
                               TRUE ~ "other")) %>% 
  ggplot(aes(x = rel_abund, y = sample, fill = top_species)) +
    geom_col() +
    theme_void() +
    scale_y_discrete(limits = day_order) +
    #my_scale2 +
    theme(axis.text = element_text())
```

### Mean proportions

The proportions of species across sample type (i.e. calculus, medium, saliva)
were calculated by taking the sum of species counts per sample
type and dividing it by the sum of all species-counts within each sample type.

The plot displays the top species across all sample types. There is little overlap
of species between sample types.

```{r}
## NOT THE WAY TO DO THIS ##
mean_proportions <- species_abundance_long %>%
  filter(str_detect(sample, "SYN")) %>% 
  rename(`#SampleID` = sample) %>%
  left_join(analysis_metadata, by = "#SampleID") %>% 
  #mutate(species = as_factor(species)) %>% 
  #rowwise(species, Env) %>%
  mutate(count = replace_na(count, 0)) %>% 
  group_by(Env, species) %>%
  summarise(sum_count = sum(count)) %>%
  group_by(Env,.add = F) %>%
  mutate(mean_prop = sum_count / sum(sum_count)) %>% 
  na.omit() %>% 
  dplyr::select(!sum_count) %>% 
  arrange(desc(mean_prop)) %>%
  slice_head(n = 6)

mean_proportions %>% 
  ggplot(aes(x = mean_prop, y = species, fill = Env)) +
    geom_col(position = "dodge", width = 0.7)
```

### Oxygen tolerance

Proportional distribution of oxygen tolerances across experimental samples.
Data on oxygen tolerance obtained from [BacDive](https://bacdive.dsmz.de/).

```{r}
no_match <- unique(filter(species_properties, is.na(`Oxygen tolerance`))$species) # which species had no match?
```


```{r}
species_properties %>% 
  filter(str_detect(sample, "SYN")) %>% 
  ggplot(aes(y = sample, x = rel_abund, fill = `Oxygen tolerance`)) +
    geom_col() +
    theme_minimal() +
    #theme(axis.text.x = element_text(angle = 90)) +
    scale_y_discrete(limits = day_order) +
    scale_fill_manual(
      values = viridisLite::turbo(n = length(levels(as.factor(species_properties$`Oxygen tolerance`)))))
```

Generally an increase of anaerobes over the course of the experiment, at the
expense of facultative anaerobes. Low proportion of aerobes throughout the
experiment.

### Amylase-binding species

Relative abundance of ABS per sample. Different treatments indicated by text colour.

```{r}
species_properties %>% 
  filter(str_detect(sample, "SYN"),
         str_detect(sample, "SYN00[1-3]", negate = T)) %>% 
  ggplot(aes(y = sample, x = rel_abund, fill = abs)) +
    geom_col() +
    theme_minimal() +
    #theme(axis.text.x = element_text(angle = 90)) +
    #scale_y_discrete(limits = day_order) +
    scale_fill_manual(
      values = viridisLite::cividis(n = 2)) +
    theme(axis.text.y = element_text(colour = analysis_colours %>%
                                       filter(str_detect(`#SampleID`, "SYN"),
                                              str_detect(`#SampleID`, "SYN00[1-3]", negate = T)) %>%
                                       arrange(`#SampleID`) %>%
                                       .$treat_col))
```

```{r}
analysis_metadata %>%
  filter(str_detect(`#SampleID`, "SYN")) %>% 
  dplyr::select(`#SampleID`, treatment)
```

Table with relative abundance of ABS in samples grouped by treatment.

```{r}
## NOT WORKING PROPERLY ##
species_properties %>%
  rename("#SampleID" = sample) %>%
  left_join(analysis_metadata, by = "#SampleID") %>%
  #mutate(abs = if_else(abs == TRUE, "yes", "no")) %>%
  filter(!is.na(treatment),
         abs == TRUE) %>%
  group_by(treatment) %>%
  summarise(abs_abund = mean(rel_abund),
            sd = sd(rel_abund))
  
```

The presence of ABS was unrelated to the presence of starch.
Although, the highest proportion of ABS was found in the control samples, which
were not exposed to starch at any point during the experiment. These results
may be irrelevant since there is no amylase in the model, except during inoculation.

# Comparative samples

Looking at abundances at species and genus level. Artificial calculus samples only
(i.e. day 24 of experiment) were compared with reference plaque and calculus samples.

## Genus composition

Top 4 most abundant genera in each sample are selected to provide a manageable
overview of sample composition.

Top 10 most abundant species are selected to provide a manageable overview.
Species outside of the top 10 are labeled 'other'.

## Species composition


### Mean proportions

The proportions of species across sample type (i.e. calculus, modern calculus,
plaque, etc.) were calculated by taking the sum of species counts per sample
type and dividing it by the sum of all species-counts within each sample type.

The plot displays the top species across all sample types. There is little overlap
of species between sample types.

```{r}
## NOT WORKING PROPERLY ##
mean_proportions <- species_abundance_long %>%
  rename(`#SampleID` = sample) %>%
  left_join(analysis_metadata, by = "#SampleID") %>% 
  #mutate(species = as_factor(species)) %>% 
  #rowwise(species, Env) %>%
  mutate(count = replace_na(count, 0)) %>% 
  group_by(Env, species) %>%
  summarise(sum_count = sum(count)) %>%
  group_by(Env,.add = F) %>%
  mutate(mean_prop = sum_count / sum(sum_count)) %>% 
  na.omit() %>% 
  select(!sum_count) %>% 
  arrange(desc(mean_prop)) %>%
  slice_head(n = 5)

mean_proportions %>% 
  ggplot(aes(x = mean_prop, y = species, fill = Env)) +
    geom_col(position = "dodge", width = 0.7)
```


### Oxygen tolerance

Proportional distribution of oxygen tolerance across artificial calculus and
comparative samples.

```{r}
species_properties %>% 
  filter(str_detect(sample, "SYN[0-9]*.[A-H]", negate = T)) %>% 
  ggplot(aes(y = sample, x = rel_abund, fill = `Oxygen tolerance`)) +
    geom_col() +
    scale_fill_viridis_d() +
    theme_minimal()
    #theme(axis.text.x = element_text(angle = 90)) +
```

```{r}
## NOT WORKING PROPERLY ##
species_properties %>%
  rename("#SampleID" = sample) %>%
  left_join(analysis_metadata, by = "#SampleID") %>%
  #mutate(abs = if_else(abs == TRUE, "yes", "no")) %>%
  #filter(abs == TRUE) %>%
  group_by(Env, `Oxygen tolerance`) %>%
  summarise(rel_abund = count / sum(count),
            sd = sd(rel_abund))
```

### Amylase-binding species

```{r}
species_properties %>% 
  filter(str_detect(sample, "SYN[0-9]*.[A-H]", negate = T)) %>% 
  ggplot(aes(y = sample, x = rel_abund, fill = abs)) +
    geom_col() +
    theme_minimal() +
    scale_fill_viridis_d()
    #theme(axis.text.x = element_text(angle = 90)) +
```

Amylase-binding species are seemingly underrepresented in the artificial calculus
samples compared to the comparative samples.

```{r}
## NOT WORKING PROPERLY ##
species_properties %>%
  rename("#SampleID" = sample) %>%
  left_join(analysis_metadata, by = "#SampleID") %>%
  #mutate(abs = if_else(abs == TRUE, "yes", "no")) %>%
  #filter(abs == TRUE) %>%
  #mutate(rel_abund = na_if(rel_abund, 0)) %>% 
  group_by(Env, abs) %>%
  summarise(mean_prop = mean(abs, na.rm = T),
            sd = sd(rel_abund)) #%>%
  group_by(Env, .add = F) %>%
  mutate(rel_abund = abund / sum(abund)) #%>%
  filter(abs == T)
```

