---
title: "FTIR analysis"
output: html_notebook
---

## Introduction

```{r}
#| label: setup
#| include: false
knitr::opts_chunk$set(message = FALSE,
                      warning = FALSE)
library(tidyverse)
library(patchwork)
library(here)
ftir_data <- read_csv(here("05-results/ftir-data_long.csv"))
grind_data <- read_csv(here("05-results/grind-data_cleaned.csv"))
ftir_metadata <- read_tsv(here("01-documentation/ftir-metadata.tsv"))
```

```{r}
#| label: setup-echo
#| eval: false

# Loading libraries and data
library(tidyverse)
library(patchwork)
library(here)
ftir_data <- read_csv(here("05-results/ftir-data_long.csv"))
grind_data <- read_csv(here("05-results/grind-data_cleaned.csv"))
ftir_metadata <- read_tsv(here("01-documentation/ftir-metadata.tsv"))
```


To determine the mineral composition and crystallisation of the artificial dental
calculus samples, we used Fourier Transform Infrared (FTIR) spectroscopy. We
compared the spectra of artificial dental calculus with spectra of archaeological
dental calculus, as well as splitting factors.

### Materials

```{r}
grind_data %>%
  filter(Sample == "Artificial calculus") %>%
  group_by(day) %>% 
  summarise(
    n = n()
    )

comb_samples <- ftir_metadata %>%
  filter(str_detect(sample, "\\+")) %>%
  separate_rows(sample, sep = "\\+") %>%
  mutate(
    sample_name = str_extract(sample, "F[0-9]+.[0-9A-D1-6]+"),
    sample = case_when(
      is.na(sample_name) ~ paste0("F", day, ".", sample),
      TRUE ~ sample
      ),
    sample_name = str_extract(sample, "F[0-9]+.[0-9A-D1-6]+")
  )
```

23 model biofilm samples from days 7, 16, 20, and 24, were analysed. Reference
samples for comparison included two modern calculus samples and an archaeological
calculus sample. The
archaeological sample was an isolated tooth from Middenbeemster, a rural Dutch
site from the 19th century. Samples
`r paste0(unique(comb_samples$sample_names), coll = ", ")`
had to be combined to
provide enough material for analysis. Only samples collected on the same day
were combined with each other. Samples from days 7 and 12 did
not produce enough material for a grind curve. They were largely composed of
organics and proteins, I would not be able to produce a splitting factor, anyway.

### Methods

Samples were ground with KBr and pressed under up to two tons of pressure.
Repeated measurements were taken and after different grinding durations, and
a grind curve was produced following Asscher et al.
[-@asscherAtomicDisorder2011].
Samples were ground and analysed up to six times (a-f) for the grind curve.
The analysis was conducted at Haifa University with a Thermo Nicolet is5
spectrometer in transmission,
at 4 cm$^{-1}$ resolution, with an average of 32 scans between wavenumbers 4000
and 400 cm$^{-1}$.
The splitting factor was calculated following Asscher et al.
[-@asscherAtomicDisorder2011],
using the full width at half maximum (FWHM) of the main peak at 1035. Grinds a-f
represent different levels of grinding. Samples denoted with an 'a' were ground
for the least amount of time and 'f' the most.

### Results

```{r}
#| label: results-setup
ftir_classic <- theme_classic()
```


```{r message=FALSE, warning=FALSE}
#| label: day-7-spectra
#| fig-cap: "Two spectra from biofilm sampled on day 7 of the experiment."
ftir_data %>%
  filter(
    day == 7,
    sample != "F7",
    sample != "F7.1A6_b"
    ) %>%
  ggplot(aes(x = wavenumber, y = abs, col = sample)) +
    geom_line() +
    theme_classic() +
      scale_x_continuous(
    n.breaks = length(as.character(seq(500, 4000, by = 500)))
    )
```


Day 7 spectra show large organic component, especially proteins (and collagen?).
The hydroxyapatite peak is present at around 1040, but the doublet around 600 is
poorly defined or absent.

```{r}
#| label: day-16-spectra
#| fig-cap: "Spectra from biofilm sampled on day 16 of the experiment."
ftir_data %>%
  filter(
    day == 16,
    is.na(grind),
    sample != "F16.2D2",
    sample != "F16.1D6"
    ) %>%
  ggplot(aes(x = wavenumber, y = abs, col = sample)) +
    geom_line() +
    theme_classic() +
    scale_x_continuous(
      n.breaks = length(as.character(seq(500, 4000, by = 500)))
    )
```

Day 16, the mineral phase has become more prominent than the organic. The
hydroxyapatite peak is now the main peak, and the doublet is well-defined.
There is still some variability in relative heights of the peaks between samples.

```{r message=FALSE, warning=FALSE}
#| label: day-24-spectra
#| fig-cap: "Two spectra from the final biofilm product sampled on day 24 of the experiment."
ftir_data %>%
  filter(
    day == 24,
    is.na(grind),
    str_detect(sample, "[_]", negate = T),
    sample != "F24.1B3",
    sample != "F24.2C3",
    sample != "F24.2D3",
    sample != "F24.2A4",
    sample != "F24.2B4"
    ) %>%
  ggplot(aes(x = wavenumber, y = abs, col = sample)) +
    geom_line() +
    theme_classic() +
    scale_x_continuous(
      n.breaks = length(as.character(seq(500, 4000, by = 500)))
    )
```


Day 24, the mineral phase continues to dominate. The main peak
is the hydroxyapatite peak at around 1039 with a well-defined doublet at 604 and
565.

```{r message=FALSE, warning=FALSE}
#| label: arch-calc-spectra
#| fig-cap: "Two spectra from a single sample of archaeological calculus with different grinding levels (c and d)."
ftir_data %>%
  filter(
    str_detect(sample, "MB11_grind_[cd]")
    ) %>%
  ggplot(aes(x = wavenumber, y = abs, col = sample)) +
    geom_line() +
    theme_classic() +
    scale_x_continuous(
      n.breaks = length(as.character(seq(500, 4000, by = 500)))
    )
```

The reference calculus is similar to the later biofilm samples. The main
difference is the a lower protein component seen as a lower Amide I peak at
around 1637.

```{r}
#| label: compare-calc-spectra
#| fig-cap: "Spectra from a sample of archaeological calculus is compared with a final product biofilm sample."
ftir_data %>%
  filter(
    str_detect(sample, "MB11_grind_c") | sample == "F24.1A3"
    ) %>%
  ggplot(aes(x = wavenumber, y = abs, col = sample)) +
    geom_line() +
    theme_classic() +
    scale_x_continuous(
      n.breaks = length(as.character(seq(500, 4000, by = 500)))
    )
```


The carbonate peak
at 1421 is not as well-developed as the reference spectrum. Protein at 1654.

The reference spectrum consists of carbonate (peaks 1418, 1458, 876), hydroxyapatite
(peaks 1036, 604, 566), and protein (1637, Amide I).

Both day 24 and modern reference resemble bone.

#### Grind curves

Splitting factors were calculated from the grinding measurements, grinding curve created
exploring crystallinity (crystal size) order, and disorder. Disorder is a steep slope
and large FWHM.

Samples are similar to fresh bone and dentin. No appreciable differences between days 16, 20,
and 24.

```{r message=FALSE, warning=FALSE}
#| grind-plots
grind_all_plot <- grind_data %>%
  group_by(day, Sample) %>%
  ggplot(aes(x = FWHM, y = IRSF, col = Sample_day, shape = Sample_day)) +
  geom_point(size = 2, alpha = 0.6) +
  geom_smooth(method = "lm", se = F) +
  theme_minimal() +
  labs(col = "Sample", shape = "Sample",
       x = "FWHM of the 1035 peak",
       y = "Splitting factor")

# isolate calculus samples to see diffs between days and the 'real deal'
grind_calc_plot <- grind_data %>%
  filter(Sample == "Artificial calculus" | Sample == "Archaeological calculus") %>%
  ggplot(aes(x = FWHM, y = IRSF, col = Sample_day, shape = Sample_day)) +
  geom_point(size = 2, alpha = 0.6) +
  geom_smooth(method = "lm", se = F) +
  labs(col = "Sample", shape = "Sample",
       x = "FWHM of the 1035 peak",
       y = "Splitting factor") +
  theme_minimal() +
  theme(
    axis.text = element_blank(),
    axis.title = element_blank(),
    axis.ticks = element_blank(),
    legend.position = "none",
    panel.border = element_rect(
      colour = "grey", 
      fill = "transparent", size = 1),
    panel.background = element_rect(fill = "white")
    ) +
  scale_colour_viridis_d(end = 0.4)
```


```{r message=FALSE, warning=FALSE}
#| label: grind-curve-inset
#| fig-cap: "(A) Grinding curves of multiple substances, and (B) calculus-only."
grind_all_plot + inset_element(grind_calc_plot, left = 0.5, bottom = 0.5, right = 1, top = 1) + 
  plot_layout(guides = "collect") + plot_annotation(tag_levels = "A")
```

## Conclusion

The artificial samples resemble both the modern reference samples and the
archaeological sample in mineral composition and crystallinity. The grinding
curves for days 16, 20, and 24 were very similar, suggesting that the CPMU
solution (introduced on day 16) may not have much of an effect.
