---
title: "FTIR analysis"
always_allow_html: true
output:
  word_document: default
  html_notebook: default
  html_document:
    df_print: paged
---

## Introduction

```{r}
#| label: setup
#| include: false
knitr::opts_chunk$set(
  message = FALSE,
  warning = FALSE)
options(ggplot2.discrete.colour = function() scale_colour_viridis_d())
library(tidyverse)
library(patchwork)
library(here)
library(plotly)
ftir_data <- read_csv(here("05-results/ftir-data_long.csv"))
grind_data <- read_csv(here("05-results/grind-data_cleaned.csv"))
ftir_metadata <- read_tsv(here("01-documentation/ftir-metadata.tsv"))
```

```{r}
#| label: setup-echo
#| eval: false

# Loading libraries and data
library(tidyverse)
library(patchwork)
library(here)
ftir_data <- read_csv(here("05-results/ftir-data_long.csv"))
grind_data <- read_csv(here("05-results/grind-data_cleaned.csv"))
ftir_metadata <- read_tsv(here("01-documentation/ftir-metadata.tsv"))
```

To determine the mineral composition and crystallisation of the artificial dental
calculus samples, we used Fourier Transform Infrared (FTIR) spectroscopy. We
compared the spectra of artificial dental calculus with spectra of archaeological
dental calculus, as well as splitting factors.

## Materials

```{r}
#| label: ftir-summ
n_samples <- ftir_metadata %>%
  filter(source == "Artificial") %>%
  group_by(day) %>% 
  summarise(
    n = n(),
    weight = mean(weight_mg, na.rm = T)
    )

knitr::kable(n_samples, digits = 2, caption = "Summary of FTIR samples used in analysis, including weight in mg.")
```

`r sum(n_samples$n)` model biofilm samples from days 7, 16, 20, and 24, were analysed.
Reference samples for comparison included two modern calculus samples and an
archaeological calculus sample. The archaeological sample was an isolated tooth
from Middenbeemster, a rural Dutch site from the 19th century. Samples
`r paste0(subset(ftir_metadata, !is.na(ftir_metadata$comb))$sample_id)`
had to be combined to provide enough material for analysis. Only samples collected
on the same day were combined with each other
(`r paste(unique(subset(ftir_metadata, !is.na(ftir_metadata$comb))$comb))`). Samples from days 7 and 12 did
not produce enough material for a grind curve. They were largely composed of
organics and proteins, i.e. would not be able to produce a splitting factor, anyway.

```{r}
#| label: grind-data-summary
grind_data %>%
  filter(Sample == "Artificial calculus") %>%
  group_by(day) %>% 
  summarise(
    n = n()
    ) %>%
  knitr::kable()
```

Table: Summary of samples used in grinding curves.

## Methods

Samples were ground with KBr and pressed under up to two tons of pressure.
Repeated measurements were taken and after different grinding durations, and
a grind curve was produced following -@asscherAtomicDisorder2011.
Samples were ground and analysed up to six times (a-f) for the grind curve.
The analysis was conducted at Haifa University with a Thermo Nicolet is5
spectrometer in transmission,
at 4 cm$^{-1}$ resolution, with an average of 32 scans between wavenumbers 4000
and 400 cm$^{-1}$.
The splitting factor was calculated following -@asscherAtomicDisorder2011,
using the full width at half maximum (FWHM) of the main peak at 1035. Grinds a-f
represent different levels of grinding. Samples denoted with an 'a' were ground
for the least amount of time and 'f' the most.

## Results

```{r}
#| label: results-setup
grind_sample_order <- c(
  "Archaeological calculus",
  "Artificial calculus day 16",
  "Artificial calculus day 20",
  "Artificial calculus day 24",
  "Archaeological bone",
  "Bone-Dentine",
  "Bone-Dentine_2",
  "Enamel",
  "Enamel_2",
  "Enamel_3")
```


```{r}
#| label: ind-spectra
#| fig-cap: "Spectra from all experiment sampling days."
#| warning: false
# day 7
ftir_data %>%
  filter(
    day == 7,
    sample != "F7",
    sample != "F7.1A6_b"
    ) %>%
  ggplot(aes(x = wavenumber, y = abs, col = sample)) +
    geom_line() +
    theme_classic() +
    scale_x_reverse()
# day 12
ftir_data %>%
  filter(
    day == 12
    #sample != "F7",
    #sample != "F7.1A6_b"
    ) %>%
  ggplot(aes(x = wavenumber, y = abs, col = sample)) +
    geom_line() +
    theme_classic() +
    scale_x_reverse()

# day 16
ftir_data %>%
  filter(
    day == 16,
    is.na(grind),
    sample != "F16.2D2",
    sample != "F16.1D6",
    sample != "F16.1B2"
    ) %>%
  ggplot(aes(x = wavenumber, y = abs, col = sample)) +
    geom_line() +
    theme_classic() +
    scale_x_reverse()

# day 24 (final product)
ftir_data %>%
  filter(
    day == 24,
    is.na(grind),
    str_detect(sample, "[_]", negate = T),
    sample != "F24.1B3",
    sample != "F24.2C3",
    sample != "F24.2D3",
    sample != "F24.2A4",
    ) %>%
  ggplot(aes(x = wavenumber, y = abs, col = sample)) +
    geom_line() +
    theme_classic() +
    scale_x_reverse()
```

### Day 7

```{r}
#| label: int-plot-day7
#| warning: false
int_plot_7 <- ftir_data %>%
  filter(
    sample == "F7.1A6"
    ) %>%
  ggplot(aes(x = wavenumber, y = abs, col = sample)) +
    geom_line() +
    theme_classic() +
    scale_x_reverse()

plotly::ggplotly(int_plot_7)
```

Day 7 spectra have large O--H and amide A absorbance bands in stretching mode around 3400 cm$^{-1}$,
as well as three marked CH~3~ and CH$_2$ stretching vibrations at 2960,
2920, and 2850 cm$^{-1}$. There is a clear amide I peak at 1650 and a less
pronounced amide II peak at 1545 cm$^{-1}$. 
In the 'fingerprint' region, C--O$_3^{2-}$ at 1400 absorbance band corresponding to the v~3~ stretching vibrations, P--O~4~ absorbance band
corresponding to the v~3~ antisymmetric stretching vibrations
at 1080 cm$^{-1}$, and minor phosphate absorption bands around 500 cm$^{-1}$ in
sample F7.1A6, but absent in sample F7.2D1. The absorption bands at 1080
cm$^{-1}$ could also be a C--N stretching mode from aliphatic amines.
The relative absorbance of O--H and Amide I and II bands are higher than the
phosphate bands, representing a relatively higher content of lipids and proteins
than inorganic content. Large variation between spectra.

### Day 12

```{r}
#| label: int-plot-day12
#| warning: false
ftir_data %>%
  filter(
    sample == "F12.1D1+F12.1D2"
    ) %>%
  ggplot(aes(x = wavenumber, y = abs, col = sample)) +
    geom_line() +
    theme_classic() +
    scale_x_reverse()

ggplotly()
```

Day 12, amide I and II continue to be the dominant peaks, and a higher ratio
of both amide and O--H to PO~4~ v~3~ absorbance bands.
Three marked CH~3~ and CH$_2$ stretching vibrations at 2960,
2920, and 2850 cm$^{-1}$. Reduced variation between two of the three spectra.

What is the absorbance band at 1080?? Is it actually  Phosphate??

### Day 16

```{r}
#| label: int-plot-day16
#| warning: false
ftir_data %>%
  filter(
    sample == "F16.2B2"
    ) %>%
  ggplot(aes(x = wavenumber, y = abs, col = sample)) +
    geom_line() +
    theme_classic() +
    scale_x_reverse()

ggplotly()
```

Day 16, the ratio of O--H and amides to PO~4~ has shifted, with
the main peak shifting to the PO~4~ v~3~ absorbance band at 1040
(except in sample F16.1A2). A well-defined PO~4~ doublet at 600 and 560 is
present. Small CO$_3^{2-}$ stretching vibrations at 870 cm$^{-1}.
Decreased variability between the spectra, with most spectra exhibiting a
higher phosphate-to-protein/lipid ratio.

### Day 24

```{r}
#| label: int-plot-day24
#| warning: false
ftir_data %>%
  filter(
    sample == "F24.1D3"
    ) %>%
  ggplot(aes(x = wavenumber, y = abs, col = sample)) +
    geom_line() +
    theme_classic() +
    scale_x_reverse()

ggplotly()
```

Day 24, large O--H and amide A absorbance bands in stretching mode around 3400 cm$^{-1}$,
as well as three minor CH~3~ and CH$_2$ stretching vibrations at 2960,
2920, and 2850 cm$^{-1}$.
Main peak of spectra is PO~4~ v~3~ at 1040
cm$^{-1}$, well-defined PO~4~ doublet at 600-550 cm$^{-1}$. Amide I band,
with small amide II and III bands. Carbonate peaks also present.
Very little variation between all of the spectra.

## Across sampling days

```{r}
#| label: all-days
#| fig-cap: "Select samples from all sampling days."
# day 6: F7.1A6
# day 12: F12.1D1+F12.1D2
# day 16: F16.2B2
# day 24: F24.1D3 
ftir_data %>%
  filter(
    sample == "F7.1A6" |
    sample == "F12.1D1+F12.1D2" |
    sample == "F16.2B2" |
    sample == "F24.1D3"
    ) %>%
  mutate(abs = case_when(sample == "F7.1A6" ~ abs + 1.35,
                         sample == "F12.1D1+F12.1D2" ~ abs + 0.9,
                         sample == "F16.2B2" ~ abs + 0.45,
                         TRUE ~ abs)) %>%
  ggplot(aes(x = wavenumber, y = abs, col = sample)) +
    geom_line() +
    theme_classic() +
    scale_x_reverse() +
    theme(axis.text.y = element_blank(),
          axis.ticks.y = element_blank())
```

The main difference in the samples across the experiment, with
increasing age, is an
increase in the main peak at around 1040 cm-1, appearance of
the phosphate doublet around 600 and 560 cm-1 between day 12
and 16, an increase
in the ratio of the Amide I/II to peaks at around 1650 and 1540 cm-1,
and a reduction of the C--H2 and C--H3 stretching vibrations.

High lipid and protein content consistent with the presence of extracellular
polysaccharides and bacteria within a matrix. Microbial DNA and RNA may be
visible from peaks around 1200--800 cm$^{-1}$ on days 7 and 12, which are
later obscured by the increasing phosphate peak at 1040 cm$^{-1}$
[@jainIsolationCharacterization2013]. The presence of water indicated by
the O--H stretch is also consistent with a biofilm, which is around 90%
water [@bergerOralBiofilms2018]. As the samples mature, the ratio of proteins
and lipids to phosphates shift from predominance of organic content to
inorganic content.

Early spectra, days 7 and 12, are also similar to collagen spectra, with the
OH absorbance band at around 3400 cm$^{-1}$, amide I, II, and III peaks at
1659, 1552, and 1240 cm$^{-1}$, respectively 
[@rohanizadehUltrastructuralStudy2005; @martinezcortizasLinkingStructural2020].

## Reference samples

```{r message=FALSE, warning=FALSE}
#| label: arch-calc-spectra
#| fig-cap: "Two spectra from a single sample of archaeological calculus with different grinding levels (c and d)."
ftir_data %>%
  filter(
    sample == "ArchDC_MB11_grind_c" | str_detect(sample, "modern_ref1")
    ) %>%
  ggplot(aes(x = wavenumber, y = abs, col = sample)) +
    geom_line() +
    theme_classic() +
    scale_x_reverse()
```

The archaeological and modern reference spectra are largely indistinguishable and
consist of an O--H absorbance band (3400 cm$^{-1}$), CH~3~ bands (3000--2900 cm$^{-1}$),
carbonate (1420, 1460, 870 cm$^{-1}$), amide I band (1650 cm$^{-1}$), and phosphates
(1040, 604, 566 cm$^{-1}$).

```{r}
#| label: compare-calc-spectra
#| warning: false
#| fig-cap: "Spectra from archaeological calculus, modern reference calculus, and artificial calculus are compared. Absorbance was shifted to allow comparison of the three spectra, so the absorbance should be interpreted as relative, not absolute."
ftir_data %>%
  filter(
    str_detect(sample, "MB11_grind_c") | sample == "F24.1A3" | 
      sample == "modern_ref1"
    ) %>%
  mutate(abs = case_when(sample == "F24.1A3" ~ abs + 0.45,
                         TRUE ~ abs)) %>% 
  ggplot(aes(x = wavenumber, y = abs, col = sample)) +
    geom_line() +
    theme_classic() +
    # scale_x_continuous(
    #   n.breaks = length(as.character(seq(500, 4000, by = 500)))
    # ) +
    scale_x_reverse() +
    theme(axis.text.y = element_blank(),
          axis.ticks.y = element_blank())
```



```{r warning=FALSE}
#| label: zoom-plot
#| fig-cap: "Plot of archaeological calculus, modern reference calculus, and artificial calculus, zoomed in on region 1800-400 cm-1."
ftir_data %>%
  filter(
    str_detect(sample, "MB11_grind_c") | sample == "F24.1A3" | 
      sample == "modern_ref1",
    wavenumber < 1800,
    ) %>%
  mutate(abs = case_when(sample == "F24.1A3" ~ abs + 0.45,
                         TRUE ~ abs)) %>% 
  ggplot(aes(x = wavenumber, y = abs, col = sample)) +
    geom_line() +
    theme_classic() +
    # scale_x_continuous(
    #   n.breaks = length(as.character(seq(500, 4000, by = 500)))
    # ) +
    scale_x_reverse() +
    theme(axis.text.y = element_blank(),
          axis.ticks.y = element_blank())
```

The biofilm samples from the end of the experiment are similar to both reference
samples. The main
difference is a lower organic component in reference samples seen as a reduced
amide I peak at around 1637 compared to the carbonate peak at around 1420, and
an absence of amide II and III. Also a reduction in CH~3~ bands at 3000-2900 cm$^{-1}$.

#### Grind curves

Splitting factors were calculated from the grinding measurements, and a
grinding curve was created to explore crystallinity (crystal size) order and
disorder. Disorder is a steep slope and large FWHM.

Samples are similar to fresh bone and dentin. No appreciable differences between
days 16, 20, and 24.

```{r message=FALSE, warning=FALSE}
#| label: grind-plots
grind_all_plot <- grind_data %>%
  mutate(
    Sample_day = factor(Sample_day, levels = grind_sample_order)) %>%
  group_by(day, Sample) %>%
    ggplot(aes(x = FWHM, y = IRSF, col = Sample_day, shape = Sample_day)) +
    geom_point(size = 2, alpha = 0.6) +
    geom_smooth(method = "lm", se = F) +
    theme_minimal() +
    labs(col = "Sample", shape = "Sample",
         x = "FWHM of the 1035 peak",
         y = "Splitting factor")

# isolate calculus samples to see diffs between days and the 'real deal'
grind_calc_plot <- grind_data %>%
  filter(Sample == "Artificial calculus" | Sample == "Archaeological calculus") %>%
  ggplot(aes(x = FWHM, y = IRSF, col = Sample_day, shape = Sample_day)) +
  geom_point(size = 2, alpha = 0.6) +
  geom_smooth(method = "lm", se = F) +
  labs(col = "Sample", shape = "Sample",
       x = "FWHM of the 1035 peak",
       y = "Splitting factor") +
  theme_minimal() +
  theme(
    axis.text = element_blank(),
    axis.title = element_blank(),
    axis.ticks = element_blank(),
    legend.position = "none",
    panel.border = element_rect(
      colour = "grey", 
      fill = "transparent", size = 1),
    panel.background = element_rect(fill = "white")
    ) +
  scale_colour_viridis_d(end = 0.4)
```


```{r message=FALSE, warning=FALSE}
#| label: fig-grind-curve-inset
#| fig-cap: "(A) Grinding curves of multiple materials; and (B) calculus-only materials, including biofilm samples from three days, and an archaeological calclulus sample."
grind_all_plot + inset_element(grind_calc_plot, left = 0.5, bottom = 0.5, right = 1, top = 1) + 
  plot_layout(guides = "collect") + plot_annotation(tag_levels = "A")
```

## Discussion

Experimental samples are indicative of biofilm growth and maturation, including
mineralisation, represented by a reduction in proteins and lipids and the increasing
intensity of peaks related to hydroxyapatite (1040 and 600-550 cm$^{-1}$) and
carbonate (1420, 1460, 870 cm$^{-1}$) over the course of the experiment.

The presence of collagen in dental calculus (and saliva) is still not clear. It
may be a result of external contamination [@mackiePreservationMetaproteome2017]
or end-products
of collagen degradation (e.g. carboxyterminal telopeptide of type I collagen, ICTP)
which may or may not be related to periodontitis
[@gursoySalivaryCollagen2013]. It is more likely that the proline-rich proteins
(secondary amine) present in the samples are attributed to the biofilm
matrix.

The artificial samples from day 24 resemble both the modern reference samples and
the archaeological sample in mineral composition and crystallinity. 
Overall, the spectra of the artificial calculus show an increase in the
inorganic component and a decrease in the organic component over time.
The grinding curves for days 16, 20, and 24 were very similar, suggesting
that the CPMU solution (introduced on day 16) may not have much of an effect.
