---
title: "FTIR analysis"
format:
  #word: default
  html:
    df-print: paged
---

## Introduction

```{r}
#| label: setup-ftir
#| include: false
knitr::opts_chunk$set(
  message = FALSE,
  warning = FALSE)
options(ggplot2.discrete.colour = function() scale_colour_viridis_d())
library(dplyr)
library(ggplot2)
library(patchwork)
library(stringr)
library(here)
library(plotly)
library(readr)
ftir_data <- read_csv(here("05-results/ftir_full-data.csv"))
ftir_grind_data <- read_csv(here("05-results/ftir_grind-data.csv"))
ftir_metadata <- read_tsv(here("01-documentation/ftir-metadata.tsv"))
source(here("02-scripts/functions.R"))
source(here("02-scripts/99_ftir-figures.R"))
```

To determine the mineral composition and level of crystallisation of the model
dental calculus samples, we used Fourier Transform Infrared (FTIR) spectroscopy.
We compared the spectra of model dental calculus with spectra of archaeological
dental calculus and used a built-in Omnic search library for mineral identification
[@weinerInfraredSpectroscopy2010]. The splitting factor of carbonate hydroxyapatite
was calculated using a macro script.

## Materials

```{r}
#| label: tbl-ftir-summ
ftir_metadata |>
  filter(!is.na(analysis_id)) |> # remove samples that weren't analysed
  group_by(source, day) |>
  summarise(
    n = n(),
    weight_mg = mean(weight_mg, na.rm = TRUE)
  ) |>
  knitr::kable(
    digits = 2, 
    col.names = c("Sample type", "Sampling day", "n", "Weight (mg)"),
    caption = "Summary of FTIR samples used in analysis, including weight in mg. Weight is the weight of the entire sample, and since only a portion of the samples were analysed, weight is not indicative of the amount of calculus analysed."
  )
```

`r nrow(filter(ftir_metadata, !is.na(analysis_id)))` model biofilm samples from days 7, 16, 20, and 24 were analysed.
Reference samples for comparison included two modern calculus samples and an
archaeological calculus sample. The archaeological sample was an isolated tooth
from Middenbeemster, a rural Dutch site from the 19th century. Some samples
had to be combined to provide enough material for analysis
(see @tbl-ftir-comb). Only samples collected
on the same day were combined. No grind curves were produced for samples from
days 7 and 12. These were largely composed of organics and proteins, and did not
form enough carbonated hydroxtapatite for analysis.

```{r}
#| label: tbl-ftir-comb
ftir_metadata %>%
  filter(str_detect(analysis_id, "\\+")) %>%
  #group_by(day) #%>%
  select(sample_id, analysis_id, day) %>%
  #mutate(comb = str_remove(comb, "_grind_a")) %>%
  distinct(sample_id, .keep_all = T)
```

Table: Summary of samples used in grinding curves.

## Methods

Samples were analysed following the method presented in @asscherAtomicDisorder2011.
A few $\mu$g of each sample were repeatedly ground together with KBr and pressed
in a 7 mm die under two tons of pressure using a Specac mini-pellet press.
Repeated measurements of the splitting factor were taken after each grind, and a
grind curve was produced following @asscherAtomicDisorder2011. Samples were
ground and analysed up to six times (a-f) for the grind curve.
Samples were analysed at the Laboratory for Sedimentary Archaeology, Haifa 
University. The analysis was conducted with a Thermo Nicolet is5 spectrometer in
transmission, at 4 cm$^{-1}$ resolution, with an average of 32 scans between
wavenumbers 4000 and 400 cm$^{-1}$.
The splitting factor was calculated following @weinerStatesPreservation1990,
by dividing the sum of the height of the absorptions at 603 cm$^{-1}$ and 567 cm$^{-1}$
in the height of the valley between them. Following @asscherAtomicDisorder2011
we plotted the IRSF against the full width at half maximum (FWHM) of the main
absorption at 1035, and compared our grinding curves to the ones produced by
@asscherAtomicDisorder2011. Grinds a-f represent different levels of grinding.
Samples denoted with an 'a' were ground for the least amount of time and 'f' the
most.

## Results

### Day 7

```{r}
#| label: fig-day7-interactive
#| warning: false

ftir_data %>%
  filter(
    analysis_id == "F7.1A6"
    ) %>%
  ggplot(aes(x = wavenumber, y = abs, col = analysis_id)) +
    geom_line() +
    theme_classic() +
    scale_x_reverse()

plotly::ggplotly()
```

Day 7 spectra have large O--H and amide A absorbance bands in stretching mode around 3400 cm$^{-1}$,
as well as three marked CH~3~ and CH$_2$ stretching vibrations at 2960,
2920, and 2850 cm$^{-1}$. There is a clear amide I peak at 1650 and a less
pronounced amide II peak at 1545 cm$^{-1}$. 
In the 'fingerprint' region, C--O$_3^{2-}$ at 1450 and 1400 absorbance bands
corresponding to the v~3~ asummetric stretching vibrations, P--O~4~ absorbance band
corresponding to the v~3~ asymmetric stretching vibrations
at 1080 cm$^{-1}$, and minor phosphate absorption bands around 500 cm$^{-1}$ in
sample F7.1A6, but absent in sample F7.2D1. The absorption bands at 1080
cm$^{-1}$, and 1040-1047 cm$^{-1}$ and minor bending absorption bands of the
phosphate doublet around 605 cm$^{-1}$ and 560 cm$^{-1}$ in sample F7.2D1, but
absent in sample F7.1A6. The absorption bands at 1080 cm$^{-1}$ could be a C--N
stretching mode from aliphatic amines, but may also come from silicate contamination
(typical quartz doublet at 797, 780).
The relative absorbance of O--H and Amide I and II bands are higher than the
phosphate bands, representing a relatively higher content of lipids and proteins
than inorganic content. Large variation between spectra.
A possible brushite absorbance band can be seen at 527
cm$^{-1}$ in sample F7.1A6 and minor absorption bands of the phosphate doublet
around 605 cm$^{-1}$ and 560 cm$^{-1}$ in sample F7.2D1, as well as what may be
minor contamination of quartz. 
The exact composition of the phosphate mineral is
hard to determine at this stage, possibly due to the presence of several phosphate
minerals including amorphous phases. The relative absorbance of O–H and Amide I
and II bands are higher than the phosphate bands, representing a relatively higher
content of lipids and proteins than inorganic content. 

### Day 12

```{r}
#| label: fig-day12-interactive
#| warning: false
ftir_data %>%
  filter(
    analysis_id == "F12.1D1+F12.1D2"
    ) %>%
  ggplot(aes(x = wavenumber, y = abs, col = analysis_id)) +
    geom_line() +
    theme_classic() +
    scale_x_reverse()

ggplotly()
```

Day 12, amide I and II continue to be the dominant peaks, and a higher ratio
of both amide and O--H to PO~4~ v~3~ absorbance bands.
Three marked CH~3~ and CH$_2$ stretching vibrations at 2960,
2920, and 2850 cm$^{-1}$. Reduced variation between two of the three spectra.

### Day 16

```{r}
#| label: fig-day16-interactive
#| warning: false

ftir_data %>%
  filter(
    analysis_id == "F16.2B2"
    ) %>%
  ggplot(aes(x = wavenumber, y = abs, col = analysis_id)) +
    geom_line() +
    theme_classic() +
    scale_x_reverse()

ggplotly()
```

Day 16, the ratio of O--H and amides to PO~4~ has shifted, with
the main peak shifting to the PO~4~ v~3~ absorbance band at 1040
(except in sample F16.1A2). A well-defined PO~4~ doublet at 600 and 560 is
present. Small CO$_3^{2-}$ asymmetric stretching at 1450 cm$^{-1}$ and 1415 cm$^{-1}$,
and stretching vibrations at 875-870 cm$^{-1}$.
Decreased variability between the spectra, with most spectra exhibiting a
higher phosphate-to-protein/lipid ratio.

### Day 24

```{r}
#| label: fig-day24-interactive
#| warning: false

ftir_data %>%
  filter(
    analysis_id == "F24.1D3"
    ) %>%
  ggplot(aes(x = wavenumber, y = abs, col = analysis_id)) +
    geom_line() +
    theme_classic() +
    scale_x_reverse()

ggplotly()
```

Day 24, large O--H and amide A absorbance bands in stretching mode around 3400 cm$^{-1}$,
as well as three minor CH~3~ and CH$_2$ stretching vibrations at 2960,
2920, and 2850 cm$^{-1}$.
Main peak of spectra is PO~4~ v~3~ at 1040
cm$^{-1}$, well-defined PO~4~ doublet at 600-550 cm$^{-1}$. Amide I band,
with small amide II and III bands. Carbonate peaks also present.
Very little variation between all of the spectra.

For day 24, although the amide I and II are still present, the main component of
the samples is the carbonate hydroxyapatite with a strong absorbance band at 1039--1040
cm$^{-1}$, well-defined PO~4~ doublet at 604/605--564/566 cm$^{-1}$, and carbonate
bands at 1413--1420 and 1445--1460 cm$^{-1}$. There is little variation between
the sample spectra  ([@fig-all-days]D).


## Across sampling days

```{r}
#| label: fig-all-days
#| fig-cap: "Select samples from all sampling days."
# day 6: F7.1A6
# day 12: F12.1D1+F12.1D2
# day 16: F16.2B2
# day 24: F24.1D3 
ftir_data %>%
  filter(
    analysis_id == "F7.1A6" |
    analysis_id == "F12.1D1+F12.1D2" |
    analysis_id == "F16.2B2" |
    analysis_id == "F24.1D3"
    ) %>%
  mutate(abs = case_when(
    analysis_id == "F7.1A6" ~ abs + 1.35,
    analysis_id == "F12.1D1+F12.1D2" ~ abs + 0.9,
    analysis_id == "F16.2B2" ~ abs + 0.45,
    TRUE ~ abs
    )) %>%
  ggplot(aes(x = wavenumber, y = abs, col = analysis_id)) +
    geom_line() +
    theme_classic() +
    scale_x_reverse() +
    theme(axis.text.y = element_blank(),
          axis.ticks.y = element_blank())
```

The main difference in the samples across the experiment, with
increasing age, is an
increase in the main PO~4~ v~3~ peak at around 1040 cm$^{-1}$, appearance of
the phosphate doublet around 600 and 560 cm$^{-1}$ between day 12
and 16, an increase in the carbonate peaks at 1458-1450 cm$^{-1}$ and 1415-1420
cm$^{-1}$, an increase
in the ratio of the Amide I/II to peaks at around 1650 and 1540 cm$^{-1}$,
and a reduction of the C--H~2~ and C--H~3~ stretching vibrations at 2960,
2920, and 2850 cm$^{-1}$.

High lipid and protein content consistent with the presence of extracellular
polysaccharides and bacteria within a matrix. Microbial DNA and RNA may be
visible from peaks around 1200--800 cm$^{-1}$ on days 7 and 12, which are
later obscured by the increasing phosphate peak at 1040 cm$^{-1}$
[@jainIsolationCharacterization2013]. The presence of water indicated by
the O--H stretch is also consistent with a biofilm, which is around 90%
water [@bergerOralBiofilms2018]. As the samples mature, the ratio of proteins
and lipids to phosphates shift from predominance of organic content to
inorganic content in the form of carbonated hydroxyapatite.

Early spectra, days 7 and 12, are also similar to collagen spectra, with the
OH absorbance band at around 3400 cm$^{-1}$, amide I, II, and III peaks at
1659, 1552, and 1240 cm$^{-1}$, respectively 
[@rohanizadehUltrastructuralStudy2005; @martinezcortizasLinkingStructural2020].

## Reference samples

```{r message=FALSE, warning=FALSE}
#| label: arch-calc-spectra
#| fig-cap: "Two spectra from a single sample of archaeological calculus with different grinding levels (c and d)."
ftir_data %>%
  filter(
    analysis_id == "ArchDC_MB11_grind_c" |
      analysis_id == "ArchDC_MB11_grind_d"
    ) %>%
  ggplot(aes(x = wavenumber, y = abs, col = analysis_id)) +
    geom_line() +
    theme_classic() +
    scale_x_reverse()
```

The archaeological and modern reference spectra are largely indistinguishable and
consist of an O--H absorbance band (3400 cm$^{-1}$), CH~3~ bands (3000--2900 cm$^{-1}$),
carbonate (1420, 1458-1450, 875-870 cm$^{-1}$), amide I band (1650 cm$^{-1}$), and phosphates
(1040, 604, 566 cm$^{-1}$).

```{r}
#| label: compare-calc-spectra
#| warning: false
#| fig-cap: "Spectra from archaeological calculus, modern reference calculus, and artificial calculus are compared. Absorbance was shifted to allow comparison of the three spectra, so the absorbance should be interpreted as relative, not absolute."
ftir_data %>%
  filter(
    str_detect(analysis_id, "MB11_grind_c") |
      analysis_id == "F24.1A3" |
      analysis_id == "modern_ref1"
    ) %>%
  mutate(abs = case_when(
    analysis_id == "F24.1A3" ~ abs + 0.45,
    TRUE ~ abs
    )) %>% 
  ggplot(aes(x = wavenumber, y = abs, col = analysis_id)) +
    geom_line() +
    theme_classic() +
    # scale_x_continuous(
    #   n.breaks = length(as.character(seq(500, 4000, by = 500)))
    # ) +
    scale_x_reverse() +
    theme(axis.text.y = element_blank(),
          axis.ticks.y = element_blank())
```



```{r warning=FALSE}
#| label: fig-zoom-plot
#| fig-cap: "Plot of archaeological calculus, modern reference calculus, and artificial calculus, zoomed in on region 1800-400 cm-1."
ftir_data %>%
  filter(
    str_detect(analysis_id, "MB11_grind_c") |
      analysis_id == "F24.1A3" | 
      analysis_id == "modern_ref1",
    wavenumber < 1800,
    ) %>%
  mutate(abs = case_when(analysis_id == "F24.1A3" ~ abs + 0.45,
                         TRUE ~ abs)) %>% 
  ggplot(aes(x = wavenumber, y = abs, col = analysis_id)) +
    geom_line() +
    theme_classic() +
    # scale_x_continuous(
    #   n.breaks = length(as.character(seq(500, 4000, by = 500)))
    # ) +
    scale_x_reverse() +
    theme(axis.text.y = element_blank(),
          axis.ticks.y = element_blank())
```

The biofilm samples from the end of the experiment are similar to both reference
samples. The main
difference is a lower organic component in reference samples seen as a reduced
amide I peak at around 1637 compared to the carbonate peak at around 1420, and
an absence of amide II and III. Also a reduction in CH~3~ bands at 3000-2900 cm$^{-1}$.

#### Grinding curves

Splitting factors of the doublet, and FWHM of the main PO~4~ peak at 1040 were
calculated from the grinding measurements, and plotted against each other to
create grinding curves to explore crystallinity (crystal size) and the order and
disorder. Disorder is a steep slope and large FWHM.

Samples were compared to the results of @asscherAtomicDisorder2011 and
@asscherVariationsAtomic2011, and the slopes
of the trend lines for our model calculus are similar to those of fresh bone and
dentin. No appreciable differences between days 16, 20, and 24. The archaeological
dental calculus does show a slightly increased slope, although more archaeological
samples are needed to show if this is consistent in archaeological calculus.

```{r}
#| label: fig-grind-plots
#| fig-cap: "Grinding curves of archaeological and model dental calculus, and enamel. The enamel is derived from published linear equations."
grind_all_plot
```


<!-- ```{r message=FALSE, warning=FALSE}
#| label: fig-grind-curve-inset
#| fig-cap: "(A) Grinding curves of multiple materials; and (B) calculus-only materials, including biofilm samples from three days, and an archaeological calclulus sample."
grind_all_plot + inset_element(grind_calc_plot, left = 0.5, bottom = 0.5, right = 1, top = 1) + 
  plot_layout(guides = "collect") + plot_annotation(tag_levels = "A")
``` -->

## Discussion

Experimental samples are indicative of biofilm growth and maturation, including
mineralisation, represented by a reduction in proteins and lipids (1460, 1420, 870)
and the increasing intensity of peaks related to carbonated hydroxyapatite (1040,
600-550 cm$^{-1}$) over the course of the experiment.

The presence of collagen in dental calculus (and saliva) is still not clear. It
may be a result of external contamination [@mackiePreservationMetaproteome2017]
or end-products
of collagen degradation (e.g. carboxyterminal telopeptide of type I collagen, ICTP)
which may or may not be related to periodontitis
[@gursoySalivaryCollagen2013]. It is more likely that the proline-rich proteins
(secondary amine) present in the samples are attributed to the biofilm
matrix.

The artificial samples from day 24 resemble both the modern reference samples and
the archaeological sample in mineral composition and crystallinity. The steeper
slope of the archaeological sample may suggest larger crystals, which is to be
expected, as the inorganic crystals expand into the space left by degrading
organic matter over time [@weinerBiologicalMaterials2010]. 
Overall, the spectra of the artificial calculus show an increase in the
inorganic component and a decrease in the organic component over time.
The grinding curves for days 16, 20, and 24 were very similar, suggesting
that the CPMU solution (introduced on day 16) may not have much of an effect.

The archaeological and modern reference spectra are largely indistinguishable and
consist of a broad O–H absorbance band (3400 cm$^{-1}$) related to hydroxyl group,
weak CH3 bands (3000–2900 cm$^{-1}$) , carbonate (1420, 1458-1450, 875-870 cm$^{-1}$),
amide I band (1650 cm$^{-1}$) which is related to the protein content, and phosphates
(1036-1040, 602-4, 563-566 cm$^{-1}$) which together withthe hydroxyl and the carbonate
can be identified as derived from carbonate hydroxyapatite, the main mineral found
in mature dental calculus (Hayashizaki et al., 2008; [@jinSupragingivalCalculus2002]).
The model
calculus samples from the end of the experiment are similar to both the modern and
archaeological reference samples. The main difference is a lower organic component
in reference samples seen as a reduced amide I peak at around 1637 compared to the
carbonate peak at around 1420, and an absence of amide II and III. Also a reduction
in CH3 bands at 3000-2900 cm$^{-1}$.
