---
title: "PCA and diversity analysis"
output: html_notebook
---

```{r setup, include=FALSE}
library(vegan)
library(mixOmics)
library(tidyverse)
library(here)
otu_table <- read_tsv(here("05-results/post-decontam_taxatable.tsv"))
metadata <- read_tsv(here("01-documentation/metadata.tsv"))
```

# Data preparation

Analysis is conducted on sample and comparative data where contaminants have
been removed (see [authentication report](./authentication.Rmd)).

The data frame with species counts is converted to the appropriate format with one
sample per row, and one column per taxon

```{r convert-table}
species_otu_table <- otu_table %>%
  column_to_rownames(var = "#OTU ID") %>%
  t()

species_otu_table <- species_abundance_long %>%
  dplyr::select(!rel_abund) %>% 
  pivot_wider(
    names_from = "species",
    values_from = "count"
    ) %>%
  mutate(across(where(is.numeric), replace_na, 0))

genus_otu_table <- genus_abundance_long %>%
  dplyr::select(!rel_abund) %>% 
  pivot_wider(
    names_from = "genus",
    values_from = "count"
    ) %>%
  mutate(across(where(is.numeric), replace_na, 0))
```

Metadata for the included samples are prepared,

```{r metadata}
analysis_metadata <- species_otu_table %>%
  #dplyr::select(sample) %>%
  #rename(`#SampleID` = sample) %>%
  left_join(metadata) # include only samples in comb_otu_table
```

and for samples from this study only.

```{r}
sample_metadata <- experiment_metadata %>%
  filter(`#SampleID` %in% analysis_metadata$`#SampleID`)
```


Data are converted to a matrix for further analysis with the
[**mixOmics**](http://mixomics.org/) and
[**vegan**](https://cran.r-project.org/web/packages/vegan/index.html)
R packages.

```{r}
species_otu_matrix <- species_otu_table

genus_otu_matrix <- genus_otu_table %>%
  column_to_rownames(var = "sample") %>%
  as.matrix()
```

# Diversity

## Alpha

### Experiment duration

Shannon index across the course of the experiment.

<!-- need to combine days into larger groups -->

```{r}
# Shannon index for all samples
alpha_div <- diversity(species_otu_matrix)

alpha_div %>% 
  as_tibble(rownames = "sample") %>%
  rename("#SampleID" = sample) %>% 
  left_join(experiment_metadata, by = "#SampleID") %>% 
  mutate(day = as_factor(day)) %>% 
  filter(!is.na(day)) %>% 
  ggplot(aes(x = day, y = value, col = day)) +
    geom_boxplot() +
    geom_jitter(width = 0.2) +
    theme_minimal() +
    theme(axis.text.x = element_text(angle = 45, vjust = 0.5, hjust = 0.4),
          legend.position = "none",
          panel.grid.major.x = element_blank())
```

Not really enough samples to make any relevant interpretations.

Days were grouped for increased sample sizes.

- inoc = days 0,3,5,7
- treatment = days 9,7,12,15,18,21
- final = day 24

```{r}
experiment_metadata_ext <- experiment_metadata %>%
  mutate(day_group = case_when(day < 8 ~ "inoc",
                               day > 8 & day < 24 ~ "treatm",
                               day == 24 ~ "final")
         )

alpha_div %>% 
  as_tibble(rownames = "sample") %>%
  rename("#SampleID" = sample) %>% 
  left_join(experiment_metadata_ext, by = "#SampleID") %>% 
  mutate(day = as_factor(day)) %>% 
  filter(!is.na(day)) %>% 
  ggplot(aes(x = day_group, y = value, col = day_group)) +
    geom_boxplot() +
    geom_jitter(width = 0.2) +
    theme_minimal() +
    theme(axis.text.x = element_text(angle = 45, vjust = 0.5, hjust = 0.4),
          legend.position = "none",
          panel.grid.major.x = element_blank())
```

Slight decrease in Shannon diversity ove the course of the experiment.

### Sample types

Shannon index compared between sample types.

```{r}
alpha_div %>%
  as_tibble(rownames = "sample") %>%
  rename("#SampleID" = sample) %>% 
  left_join(analysis_metadata, by = "#SampleID") %>% 
  filter(Env != "sediment", 
         Env != "skin",  
         Env != "stool", 
         Env != "indoor_air") %>%
  remove_missing() %>%
  ggplot(aes(x = Env, y = value, col = Env)) +
    geom_boxplot() +
    geom_jitter(width = 0.2) +
    theme_minimal() +
    theme(axis.text.x = element_text(hjust = 1, angle = 45),
          legend.position = "none",
          panel.grid.major.x = element_blank())
```

Higher overall alpha diversity in the comparative samples than the artificial
samples, both medium and final calculus product.


## Beta

sPCA using the **mixOmics** R package.

Centered log-ratio (CLR) transformation with 1 offset.

### Experiment samples only

```{r}
clr_sample <- logratio.transfo(sample_otu_matrix, "CLR", 1)
```

```{r}
spca_sample <- spca(clr_sample, ncomp = 10, scale = F)
sample_explain_var <- spca_sample$prop_expl_var$X
```

```{r}
plot(spca_sample, type = "l")
```


```{r}
sample_princomp <- spca_sample$x %>%
  as_tibble(rownames = "sample")
sample_princomp %>%
  ggplot(
    aes(
      x = PC1, y = PC2, 
      col = as.factor(sample_metadata$day),
      shape = sample_metadata$Env
     )
    ) +
    geom_point(size = 4) +
    labs(
      x = paste(
        "PC1", 
        scales::percent(sample_explain_var[[1]], accuracy = 0.1)),
      y = paste(
        "PC2", 
        scales::percent(sample_explain_var[[2]], accuracy = 0.1)),
      col = "Sampling day",
      shape = "Sample type"
      ) +
    theme_bw() +
    scale_shape_discrete(solid = F)
    scale_colour_viridis_d()
```

PC1 discriminates the saliva samples from the medium and calculus samples,
while PC2 mainly discriminates calculus from medium.

### Artificial calculus and comparative oral sources

```{r}
comp_otu_matrix <- species_otu_matrix %>%
  as_tibble(rownames = "sample") %>%
  filter(sample %in% analysis_metadata$`#SampleID`) %>%
  column_to_rownames(var = "sample")

clr_species <- logratio.transfo(comp_otu_matrix, "CLR", 1)
clr_genus <- logratio.transfo(genus_otu_matrix, "CLR", 1)
```

Scree plot of components.

```{r}
spca_species <- spca(clr_species, ncomp = 10, scale = F)

spca_genus <- spca(clr_genus, ncomp = 10, scale = F)

plot(spca_species)

plot(spca_genus)
```

sPCA plot

```{r}
spca_compar <- spca_species$x %>%
  as_tibble(rownames = "sample")
compar_explain_var <- spca_species$prop_expl_var$X

spca_compar %>%
  ggplot(aes(x = PC1, y = PC2, col = analysis_metadata %>%
               filter(`#SampleID` %in% rownames(comp_otu_matrix)) %>%
               .$Env)) +
    geom_point(size = 2) +
    theme_bw() +
    labs(x = paste(
      "PC1", scales::percent(compar_explain_var[[1]], accuracy = 0.1)
      ),
      y = paste(
        "PC2", scales::percent(compar_explain_var[[2]], accuracy = 0.1)
      )
    ) +
  scale_colour_viridis_d() +
  labs(col = "Env")
```

The artificial calculus clusters with the medium, while the modern
reference calculus clusters with plaque and saliva.

Main drivers of discrimination. (include loading vectors)

Top loading vectors for PC1 and PC2.

```{r}
spca_species$rotation %>%
  as_tibble(rownames = "species") %>% 
  arrange(desc(PC1))
```

```{r}
plotVar(spca_species, cutoff = 0.92)
```

```{r}
biplot(spca_species, comp = 1:2, cutoff = 0.92, group = analysis_metadata$Env, ind.names = F)
```


```{r}
plotVar(spca_genus, cutoff = 0.9)
```


```{r}
biplot(spca_genus, comp = 1:2, cutoff = 0.92, group = analysis_metadata$Env, ind.names = F)
```


